---
title: "Statistical Inference - Project Part 1"
author: "Yakov Falkovich"
date: "Wednesday, September 09, 2015"
output: pdf_document
---
# Simulation
I'm simulating the mean of 40 exponentially distributed variables with $\lambda=0.2$ 1000 times:
```{r}
set.seed(1)
meansim = sapply(c(1:1000), function(i) {mean(rexp(40,0.2))})
```

The result of the simulation looks like:
```{r,echo=FALSE} 
head(meansim) 
```

The expected mean of this sample is simply the mean of an exponential variable, $1/\lambda=5$. The sample mean is 
```{r} 
round(mean(meansim),2)
```

A single exponential variable has variance $\sigma^2 = 1/\lambda^2$ so a mean of $N=40$ variables should have variance $\sigma^2/N = \frac{1}{\lambda^2 \cdot N} = \frac{1}{0.2^2 \cdot 40} = 0.625$. The variance of the sample is in fact is 
```{r}
round(var(meansim),2)
```

Both parameters are in fact very close to their expected values.  

---
#Testing for Normality
Let's compare the distributions of 1000 means of 40 exponential variables with 1000 unaveraged exponential variables.

First we normalize the results by subtracting the expected mean and dividing by the expected standard 
```{r}
meannorm=(meansim-5)/(5/sqrt(40))
```
```{r, echo = FALSE, fig.width=5, fig.height = 6, fig.align='center'}
library(ggplot2)
q1<-qplot(meannorm, geom = 'blank')+geom_histogram(aes(x=meannorm,y=..density..),binwidth=.3,fill="cornflowerblue",color="black")+stat_function(fun = dnorm, size = 1.5,color="green4")
print(q1)
```
Compare that to 1000 samples of a normalized exponential variable:
```{r}
expsim=rexp(1000,0.2)
```
```{r, echo = FALSE, fig.width=5, fig.height = 6, fig.align='center'}
dexp = function(x) {0.2*exp(-x*0.2)}
q2<-qplot(expsim, geom = 'blank')+geom_histogram(aes(x=expsim,y=..density..),binwidth=2,fill="brown1",color="black")+stat_function(fun = dexp, size = 1.5,color="green4")
print(q2)
```